State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-11 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
moved LS appendix to the main body

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state vector sequence estimate (see~\ref{App:LeastSquaresAppendix}).

% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
% Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearising about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, while maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, since the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to high computation demands.

@@

\section{Parameter Estimation Using Least Squares}\label{App:LeastSquaresAppendix} 
The finite state space model given in \ref{eq:finalformstatespacemodel} is linear with respect to parameters. Therefore a least square (LS) method can be applied to estimate the unknown parameters. To create the least squares parameter estimator we first define the $L \times n_{\theta}$ matrix
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
Now given a state estimate sequence from the unscented Rauch Tung Striebel smoother (URTSS), we can write
	\mathbf x_{1} &=& \mathbf{q}(\mathbf x_0) \boldsymbol{\theta}+\xi\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2} &=& \mathbf{q}(\mathbf x_1) \boldsymbol{\theta}+\xi\mathbf x_1+\mathbf e_1 \nonumber \\
	&\vdots& \\
	\mathbf x_{T}&=&\mathbf{q}(\mathbf x_{T-1}) \boldsymbol{\theta}+\xi\mathbf x_{T-1}+\mathbf e_{T-1}. \nonumber \\
This can be written in the compact form
	\mathbf Z=\mathbf X \mathbf W+\mathbf{e} 
	\mathbf Z=\left[
	\begin{array}{cccc}
		\mathbf x_{1}\\
		\mathbf x_{2}\\
		\vdots\\
		\mathbf x_{T}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\mathbf x_0)& \mathbf x_{0}\\
		\mathbf q(\mathbf x_1)& \mathbf x_{1}\\
		\vdots & \vdots\\
		\mathbf q(\mathbf x_{T-1})& \mathbf x_{T-1}
and
\quad \mathbf W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\right].
Following this, the parameters, $ \mathbf W$, can be estimated by
	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\section{Parameters} 
\begin{tabular}
	{c|c} Symbol & Description \\
	\hline
	\\ & \textbf{Domain and indicies} \\
	\hline
	$\Omega$ & spatial domain \\
	$t$ & time (seconds) \\
	$\mathbf{r}$ & spatial location \\
	$n$ & sensor index $n=1,..,N$ \\
	$T$ & time step \\
	\\ & \textbf{Spatiotemporal Signals} \\
	\hline
	$y(\mathbf{r}_n,t)$ & observation \\
	$v(\mathbf{r},t)$ & mean membrane potential \\
	$g(\mathbf{r},t)$ & average action potential rate \\
	$f(\mathbf{r},t)$ & firing function rate \\
	$u(\mathbf{r},t)$ & external input \\
	$e(\mathbf{r},t)$ & field disturbance, covariance $\gamma$\\
	$\epsilon(\mathbf{r}_n,t)$ & observation noise, covariance $\Sigma_\epsilon$ \\
	\\ & \textbf{Model} \\
	\hline
	$h(t)$ & post-synaptic response kernel \\
	$m(\mathbf{r},\mathbf{r}')$ & sensor kernel, variance $\sigma_m^2$ \\
	$\eta(t)$ & Heaviside function \\
	$\zeta$ & inverse synaptic time constant \\
	$w(\mathbf{r},\mathbf{r}')$ & spatial connectivity kernel \\
	$f_{max}$ & maximal firing rate \\
	$\varsigma$ & slope of sigmoidal activation function \\
	$v_0$ & firing threshold \\
	$\delta(t)$ & Dirac-delta function \\
	$D$ & Temporal differential operator \\
	$\xi$ & time constant parameter \\
	\\ & \textbf{Decomposition} \\
	\hline
	$L$ & number of field basis functions  \\
	$\mathbf{\phi(r)}$ & vector of Gaussian basis functions \\
	$\mathbf{x}_t$ & state vector at time $t$ \\
	$\mathbf{\psi}$ & vector of connectivity kernel basis functions \\
	$\theta$ & vector of connectivity kernel parameters \\
	$\Gamma$ & inner product of field basis functions \\
	$q()$ & state function \\
	$k()$ & maps into to discrete state of input function \\
	$\mathbf{e}_t$ & state disturbance, covariance $\Sigma_e$ \\
	$\mathbf{C}$ & observation matrix \\
	\\ & \textbf{Frequency Analysis} \\
	\hline
	$V(\nu)$ & Spectrum of the dynamic field \\
	$\nu$ & Spatial frequency \\
	$\nu_c$ & Spatial cut-off frequency \\
	$\Delta_s$ & Distance between adjacent sensors \\
	$\rho$ & Over-sampling parameter \\
	$\Delta_b$ & Distance between field basis functions \\
	$\sigma_{nu}^2$ & Variance of FT of Gaussian basis function \\
	$\sigma^2$ & Spatial variance of Gaussian in spatial domain \\
	\\ & \textbf{State Estimation} \\
	\hline
	$\hat{\mathbf{x}}$ & State estimate \\
	$\chi$ & Matrix of sigma vectors \\
	$\hat{\mathbf{x}}_t^{f-}$ & Forward prior estimate \\
	$\hat{\mathbf{x}}_t^f$ & Forward posterior estimate \\
	$\hat{\mathbf{x}}_t^{b-}$ & Backward prior estimate \\
	$\hat{\mathbf{x}}_t^{b}$ & Backward posterior estimate \\
	$P^f_t$ & Forward posterior covariance matrix \\
	$P^{f-}_t$ & Forward prior covariance matrix \\
\end{tabular}
\section*{References} 
\bibliographystyle{unsrt} 
\bibliography{BrainIDE}

>>>>

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state vector sequence estimate. To create the least squares parameter estimator we first define the $L \times n_{\theta}$ matrix
\begin{equation}
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'.
\end{equation}
Now given a state estimate sequence from the unscented Rauch Tung Striebel smoother (URTSS), we can write
\begin{eqnarray}
	\mathbf x_{1} &=& \mathbf{q}(\mathbf x_0) \boldsymbol{\theta}+\xi\mathbf x_0+\mathbf e_0 \nonumber \\
	\mathbf x_{2} &=& \mathbf{q}(\mathbf x_1) \boldsymbol{\theta}+\xi\mathbf x_1+\mathbf e_1 \nonumber \\
	&\vdots& \\
	\mathbf x_{T}&=&\mathbf{q}(\mathbf x_{T-1}) \boldsymbol{\theta}+\xi\mathbf x_{T-1}+\mathbf e_{T-1}. \nonumber \\
\end{eqnarray}
This can be written in the compact form
\begin{equation}
	\mathbf Z=\mathbf X \mathbf W+\mathbf{e} 
\end{equation}
where
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\mathbf x_{1}\\
		\mathbf x_{2}\\
		\vdots\\
		\mathbf x_{T}
	\end{array}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\mathbf x_0)& \mathbf x_{0}\\
		\mathbf q(\mathbf x_1)& \mathbf x_{1}\\
		\vdots & \vdots\\
		\mathbf q(\mathbf x_{T-1})& \mathbf x_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathbf W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\end{array}
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\end{array}
	\right].
\end{equation*}
\end{small}
Following this, the parameters, $ \mathbf W$, can be estimated by
\begin{equation}
	\mathbf W\approx(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\end{equation}
% The accuracy of the approximation is linked to the time step used in the temporal discretisation of the system, where the state transition appears more linear with a finer resolution. 
% Alternatives to the UKF are the Extended Kalman Filter (EKF)~\cite{Haykin2001} and the Sequential Monte Carlo (SMC) filter~\cite{doucet2001}. The EKF approximates the state transition equation by linearising about the current state estimates (by calculating the Jacobian) at each time instance. The linearisation maintains the Gaussianality of the model. The unscented Kalman filter approximates the posterior state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, while maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, since the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the EKF can be problematic. SMC filtering can theoretically provide an exact posterior state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to high computation demands.

@@

\section{Parameters} 
\begin{tabular}
	{c|c} Symbol & Description \\
	\hline
	\\ & \textbf{Domain and indicies} \\
	\hline
	$\Omega$ & spatial domain \\
	$t$ & time (seconds) \\
	$\mathbf{r}$ & spatial location \\
	$n$ & sensor index $n=1,..,N$ \\
	$T$ & time step \\
	\\ & \textbf{Spatiotemporal Signals} \\
	\hline
	$y(\mathbf{r}_n,t)$ & observation \\
	$v(\mathbf{r},t)$ & mean membrane potential \\
	$g(\mathbf{r},t)$ & average action potential rate \\
	$f(\mathbf{r},t)$ & firing function rate \\
	$u(\mathbf{r},t)$ & external input \\
	$e(\mathbf{r},t)$ & field disturbance, covariance $\gamma$\\
	$\epsilon(\mathbf{r}_n,t)$ & observation noise, covariance $\Sigma_\epsilon$ \\
	\\ & \textbf{Model} \\
	\hline
	$h(t)$ & post-synaptic response kernel \\
	$m(\mathbf{r},\mathbf{r}')$ & sensor kernel, variance $\sigma_m^2$ \\
	$\eta(t)$ & Heaviside function \\
	$\zeta$ & inverse synaptic time constant \\
	$w(\mathbf{r},\mathbf{r}')$ & spatial connectivity kernel \\
	$f_{max}$ & maximal firing rate \\
	$\varsigma$ & slope of sigmoidal activation function \\
	$v_0$ & firing threshold \\
	$\delta(t)$ & Dirac-delta function \\
	$D$ & Temporal differential operator \\
	$\xi$ & time constant parameter \\
	\\ & \textbf{Decomposition} \\
	\hline
	$L$ & number of field basis functions  \\
	$\mathbf{\phi(r)}$ & vector of Gaussian basis functions \\
	$\mathbf{x}_t$ & state vector at time $t$ \\
	$\mathbf{\psi}$ & vector of connectivity kernel basis functions \\
	$\theta$ & vector of connectivity kernel parameters \\
	$\Gamma$ & inner product of field basis functions \\
	$q()$ & state function \\
	$k()$ & maps into to discrete state of input function \\
	$\mathbf{e}_t$ & state disturbance, covariance $\Sigma_e$ \\
	$\mathbf{C}$ & observation matrix \\
	\\ & \textbf{Frequency Analysis} \\
	\hline
	$V(\nu)$ & Spectrum of the dynamic field \\
	$\nu$ & Spatial frequency \\
	$\nu_c$ & Spatial cut-off frequency \\
	$\Delta_s$ & Distance between adjacent sensors \\
	$\rho$ & Over-sampling parameter \\
	$\Delta_b$ & Distance between field basis functions \\
	$\sigma_{nu}^2$ & Variance of FT of Gaussian basis function \\
	$\sigma^2$ & Spatial variance of Gaussian in spatial domain \\
	\\ & \textbf{State Estimation} \\
	\hline
	$\hat{\mathbf{x}}$ & State estimate \\
	$\chi$ & Matrix of sigma vectors \\
	$\hat{\mathbf{x}}_t^{f-}$ & Forward prior estimate \\
	$\hat{\mathbf{x}}_t^f$ & Forward posterior estimate \\
	$\hat{\mathbf{x}}_t^{b-}$ & Backward prior estimate \\
	$\hat{\mathbf{x}}_t^{b}$ & Backward posterior estimate \\
	$P^f_t$ & Forward posterior covariance matrix \\
	$P^{f-}_t$ & Forward prior covariance matrix \\
\end{tabular}
\section*{References} 
\bibliographystyle{unsrt} 
\bibliography{BrainIDE}
